Escape Analysis and Specialization in a Tracing JIT

### 2010-09-28 talk/pepm2011/paper.tex
convert all traces to use math style.

\begin{figure}
\begin{verbatim}
    # arguments to the trace: p0, p1
    # inside f: res.add(y)
    guard_class(p1, BoxedInteger)
        # inside BoxedInteger.add
        i2 = getfield_gc(p1, intval)
        guard_class(p0, BoxedInteger)
            # inside BoxedInteger.add__int
            i3 = getfield_gc(p0, intval)
            i4 = int_add(i2, i3)
            p5 = new(BoxedInteger)
                # inside BoxedInteger.__init__
                setfield_gc(p5, i4, intval)
    # inside f: BoxedInteger(-100) 
    p6 = new(BoxedInteger)
        # inside BoxedInteger.__init__
        setfield_gc(p6, -100, intval)

    # inside f: .add(BoxedInteger(-100))
    guard_class(p5, BoxedInteger)
        # inside BoxedInteger.add
        i7 = getfield_gc(p5, intval)
        guard_class(p6, BoxedInteger)
            # inside BoxedInteger.add__int
            i8 = getfield_gc(p6, intval)
            i9 = int_add(i7, i8)
            p10 = new(BoxedInteger)
                # inside BoxedInteger.__init__
                setfield_gc(p10, i9, intval)

    # inside f: BoxedInteger(-1)
    p11 = new(BoxedInteger)
        # inside BoxedInteger.__init__
        setfield_gc(p11, -1, intval)

    # inside f: y.add(BoxedInteger(-1))
    guard_class(p0, BoxedInteger)
        # inside BoxedInteger.add
        i12 = getfield_gc(p0, intval)
        guard_class(p11, BoxedInteger)
            # inside BoxedInteger.add__int
            i13 = getfield_gc(p11, intval)
            i14 = int_add(i12, i13)
            p15 = new(BoxedInteger)
                # inside BoxedInteger.__init__
                setfield_gc(p15, i14, intval)

    # inside f: y.is_positive()
    guard_class(p15, BoxedInteger)
        # inside BoxedInteger.is_positive
        i16 = getfield_gc(p15, intval)
        i17 = int_gt(i16, 0)
    # inside f
    guard_true(i17)
    jump(p15, p10)
\end{verbatim}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new}, \texttt{set/getfield\_gc} and
\texttt{guard\_class} operations.

@@

XXX simplify traces a bit more
get rid of \_gc suffix in set/getfield\_gc

@@

The objects that are allocated in the example trace in
Figure~\ref{fig:unopt-trace} fall into categories 1 and 3. Objects stored in
\texttt{p5, p6, p11 XXX} are in category 1, objects in \texttt{p10, p15} are in
category 3.

@@

\begin{verbatim}
p5 = new(BoxedInteger)
setfield_gc(p5, i4, intval)
p6 = new(BoxedInteger)
setfield_gc(p6, -100, intval)
\end{verbatim}

The virtual object stored in \texttt{p5} would know that it is an \texttt{BoxedInteger}, and that
the \texttt{intval} field contains \texttt{i4}, the one stored in \texttt{p6} would know that
its \texttt{intval} field contains the constant -100.

The following operations, that use \texttt{p5} and \texttt{p6} could then be
optimized using that knowledge:

\begin{verbatim}
guard_class(p5, BoxedInteger)
i7 = getfield_gc(p5, intval)
# inside BoxedInteger.add
guard_class(p6, BoxedInteger)
# inside BoxedInteger.add__int
i8 = getfield_gc(p6, intval)
i9 = int_add(i7, i8)
\end{verbatim}

The \texttt{guard\_class} operations can be removed, because the classes of \texttt{p5} and
\texttt{p6} are known to be \texttt{BoxedInteger}. The \texttt{getfield\_gc} operations can be removed
and \texttt{i7} and \texttt{i8} are just replaced by \texttt{i4} and -100. Thus the only
remaining operation in the optimized trace would be:

\begin{verbatim}
i9 = int_add(i4, -100)
\end{verbatim}

@@

XXX should the variables be written in $math-style$ everywhere?

\texttt{
\begin{tabular}{l} 
$p_{15}$ = new(BoxedInteger) \\
setfield\_gc($p_{15}$, $i_{14}$, intval) \\
$p_{10}$ = new(BoxedInteger) \\
setfield\_gc($p_{10}$, $i_{9}$, intval) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the forced virtual objects have been removed (some \texttt{getfield\_gc} operations
and \texttt{guard\_class} operations).

@@

The two new \texttt{BoxedIntegers} stored in \texttt{p15} and \texttt{p10} are passed into
the next iteration of the loop. The next iteration will check that they are
indeed \texttt{BoxedIntegers}, read their \texttt{intval} fields and then not use them
any more. Thus those instances are in category 3.

>>>>

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = getfield($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = getfield($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{5}$, $i_{4}$, intval) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{6}$, -100, intval) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = getfield($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = getfield($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{10}$, $i_{9}$, intval) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{11}$, -1, intval) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = getfield($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = getfield($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{15}$, $i_{14}$, intval) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = getfield($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new}, \texttt{set/getfield} and
\texttt{guard\_class} operations.

@@

XXX simplify traces a bit more

@@

The objects that are allocated in the example trace in
Figure~\ref{fig:unopt-trace} fall into categories 1 and 3. Objects stored in
$p_{5}$, $p_{6}$, $p_{11}$ are in category 1, objects in $p_{10}$, $p_{15}$ are in
category 3.

@@

\texttt{
\begin{tabular}{l} 
$p_{5}$ = new(BoxedInteger) \\
setfield($p_{5}$, $i_{4}$, intval) \\
$p_{6}$ = new(BoxedInteger) \\
setfield($p_{6}$, -100, intval) \\
\end{tabular}
}

The virtual object stored in $p_{5}$ would know that it is an \texttt{BoxedInteger}, and that
the \texttt{intval} field contains $i_{4}$, the one stored in $p_{6}$ would know that
its \texttt{intval} field contains the constant -100.

The following operations, that use $p_{5}$ and $p_{6}$ could then be
optimized using that knowledge:

\texttt{
\begin{tabular}{l} 
guard\_class($p_{5}$, BoxedInteger) \\
$i_{7}$ = getfield($p_{5}$, intval) \\
\# inside BoxedInteger.add \\
guard\_class($p_{6}$, BoxedInteger) \\
\# inside BoxedInteger.add\_\_int \\
$i_{8}$ = getfield($p_{6}$, intval) \\
$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
\end{tabular}
}

The \texttt{guard\_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{getfield} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

\texttt{
\begin{tabular}{l} 
$i_{9}$ = int\_add($i_{4}$, -100) \\
\end{tabular}
}

@@

\texttt{
\begin{tabular}{l} 
$p_{15}$ = new(BoxedInteger) \\
setfield($p_{15}$, $i_{14}$, intval) \\
$p_{10}$ = new(BoxedInteger) \\
setfield($p_{10}$, $i_{9}$, intval) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the forced virtual objects have been removed (some \texttt{getfield} operations
and \texttt{guard\_class} operations).

@@

The two new \texttt{BoxedIntegers} stored in $p_{15}$ and $p_{10}$ are passed into
the next iteration of the loop. The next iteration will check that they are
indeed \texttt{BoxedIntegers}, read their \texttt{intval} fields and then not use them
any more. Thus those instances are in category 3.
